---
title: "Introduction to rasterly"
author: "Zehao Xu"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to rasterly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{grid}
---
  
```{r library, eval = TRUE, echo = TRUE, fig.align="center", fig.width = 6, fig.height = 4, out.width = "75%", warning=FALSE, message=FALSE}
library(rasterly)
library(data.table)
library(lubridate)
library(grid)
library(plotly)
```

`rasterly` is a package to display large data (million or billion) which generally cannot be handled by `R` graphics systems. The idea of `rasterly` comes from python library `datashader` http://datashader.org/getting_started/index.html. The speed of `rasterly` is comparable with `datashader` and the code will be optimized in the future. However, `rasterly` provides flexible operating interface, like `ggplot2` mapping system `aes()`, combining different elements into layers and etc

## Data set

The data set is highlights Uber trips taken in New York City from April 1 2014 to September 30 2014.

```{r data}
# Load data
ridesRaw_1 <- "https://raw.githubusercontent.com/plotly/datasets/master/uber-rides-data1.csv" %>%
  data.table::fread(stringsAsFactors = FALSE)
ridesRaw_2 <- "https://raw.githubusercontent.com/plotly/datasets/master/uber-rides-data2.csv" %>% 
  data.table::fread(stringsAsFactors = FALSE)
ridesRaw_3 <- "https://raw.githubusercontent.com/plotly/datasets/master/uber-rides-data3.csv"  %>% 
  data.table::fread(stringsAsFactors = FALSE)
ridesDf <- list(ridesRaw_1, ridesRaw_2, ridesRaw_3) %>% 
  data.table::rbindlist()

# get hour
time <- lubridate::ymd_hms(ridesDf$`Date/Time`)
ridesDf <-  ridesDf[, 'Date/Time':=NULL][, list(Lat, 
                                                Lon,
                                                hour = lubridate::hour(time), 
                                                month = lubridate::month(time),
                                                day = lubridate::day(time))]
head(ridesDf)
```

This data set has 4533327 observations, containing "latitude", "longitude", "hour", "month" and "day". 

## Basic

It may take a couple of minutes by basic `graphics::plot()`. How about "rasterizing" the image?

```{r basic, warning=FALSE, message=FALSE, fig.width = 4, fig.height = 3}
start_time <- Sys.time()
p <- ridesDf %>% 
  rasterly(mapping = aes(x = Lat, y = Lon)) %>% 
  rasterize_points()
p
end_time <- Sys.time()
end_time - start_time
```

## `rasterly` Structure

#### Subsetting

`rasterly()` is a parent layer for initial settings, which includes `plot_height`, `plot_width` and others for generating rasters; then, children layers such as `rasterize_points()` can be piped into. Note that, "p" is a list of environments.

```{r list return}
# A list of environments
str(p)
```

The elements in "p" can be easily extracted or replaced by `[` and `[<-`

```{r subsetting, warning=FALSE, message=FALSE, fig.width = 4, fig.height = 3}
p["background"]
# background in children layer `rasterly_points()` is replaced
p["background", level = 2] <- "black"
p["background"]
# colour_map in both `rasterly()` and `rasterly_points()` are replaced
## fire is a colour vector with length 256
## see `rasterly::fire`
p["colour_map", level = 1:2] <- fire
p
```

  * `level` helps to define which layer to replace and the default setting is 1 (the parent layer `rasterly()`).
  * Available states can be extracted or replaced are listed here:
    1. Aggregation: `data`, `mapping`, `plot_width`, `plot_height`, `range`, `x_range`, `y_range`, `xlim`, `ylim`, `aesthetics`, `reduction_func`, `glyph`, `max_size`, `group_by_data_table`, `drop_data`, `variable_check`
    2. Display: `background`, `colour_map`, `colour_key`, `alpha`, `span`, `show_raster`, `layout`
    
#### Build rasterly by `rasterly_build()`

Display info can be achieved via `rasterly_build()`

```{r rasterly_build}
build <- rasterly_build(p)
str(build)
```

It contains:

  * agg: aggregation matrices, a list of numerical matrices
  * image: a raster matrix (has the same dimension with aggregation matrices)
  * lims: a list of x limits and y limits for each layer
  * x_range: the range of x over all layers
  * y_range: the range of y over all layers
  * plot_height: plot height, aggregation matrix number of rows
  * plot_width: plot width, aggregation matrix number of columns
  * variable_names: variables names
  * background: background colour
  * colours: colour_map (colour used to map in each pixel) or colour_key (used for categorical variable. In general, 'colour_key' would be called when "colour" is set in `aes()`)

## Display

`rasterly` does not provide any display tools. It replies on the third graphics parties

#### `grid` graphics

```{r grid.raster, fig.width = 4, fig.height = 3, eval=FALSE}
grid::grid.raster(build$image)
```

`grid.rasterly()` is created based on `grid::grid.raster()` which gives labels, axis, fonts and etc to rich the display

```{r, eval=FALSE}
p # which is equivalent to
print(p) # which is equivalent to
grid.rasterly(p) # which is equivalent to
plot(p)
```

#### `plotly` graphics

  + `add_rasterly()`: "Plotly" pipes `add_trace(...)` to layer up. `rasterly` builds `add_rasterly()` from `add_heatmap()` to enable its functionality in "plotly"
    
    ```{r add_rasterly, fig.width = 4, fig.height = 3, eval=FALSE}
    plotly::plot_ly(ridesDf, x = ~Lat, y = ~Lon) %>%
      add_rasterly() %>% 
      layout(
        title = "Uber drives",
        x <- list(
          title = "Lat"
        ),
        y <- list(
          title = "Lon"
        )
      )
    ```
  
  + `plotly.rasterly()`: `plotly.rasterly` takes a `rasterly` object and returns a "plotly" object
    
    ```{r plotly_rasterly, warning=FALSE, message=FALSE, fig.width = 4, fig.height = 3, eval=FALSE}
    # plotly
    ply <- p %>% 
      plotly.rasterly(sizing = "contain")
    ply
    ```

## API

`rasterly` application programming interface

```{r}
r <- rasterly(data = ridesDf, 
              mapping = aes(x = Lat, y = Lon))
```

#### Mapping system 

  * Set `colour`

```{r set colour, fig.width = 4, fig.height = 3}
r %>% 
  rasterize_points(
     mapping = aes(colour = hour),
     colour_key = hourColours,
     background = "black"
  ) -> g
g
```

Different colours represent different hours

```{r legend, fig.width = 4, fig.height = 3}
# A legend may be added in rasterly in the furture
plot(1:24, y = rep(1,24), col = hourColours, pch = 19, cex = 3)
```

The number of aggregation matrices is equivalent to the number of categories.

```{r number of aggregation matrices}
build_g <- rasterly_build(g)
# set "1" is because we just has one layer
length(build_g$agg[[1]])
```

Colours in "image" in `build_g` is built based on weighted arithmetic mean (default) on aggregation matrices. Also, we can choose "cover" layout to display multiple aggregation matrices.

```{r set colour cover, fig.width = 4, fig.height = 3, eval=FALSE, eval=FALSE}
r %>% 
  rasterize_points(
     mapping = aes(colour = hour),
     colour_key = hourColours,
     background = "black",
     layout = "cover"
  )
```

The afterwards raster matrix will cover on previous one.

  * Set `on`
  
  `reduction_func` is implemented `on` which variable

```{r set on, fig.width = 4, fig.height = 3, eval=FALSE}
r %>% 
  rasterize_points(
    reduction_func = "mean", # take the "mean" reduction function
    mapping = aes(on = -Lat)
  )  
```

  * Set `size`
  
  Proportional size on corresponding variable. `max_size` is the largest bins shared by each observation.
  
```{r set size, fig.width = 4, fig.height = 3, eval=FALSE}
r %>% 
  rasterize_points(
    mapping = aes(size = month),
    max_size = 4
  )  
```

Currently, only `x`, `y`, `colour`, `on` and `size` can be set in mapping `aes()`.

#### Reduction function

A reduction operator function is used when aggregating data points into a each bin. 

  * `mean` reduction function:

```{r reduction on mean, fig.width = 4, fig.height = 3, eval=FALSE}
r %>% 
  rasterize_points(
    reduction_func = "mean", # take the "mean" reduction function
    background = "black",    # change background to "black"  from right to left (from dark to light)
    colour_map = fire # set a different colour_map
  )  
```

Reduction function `mean` averages the y column (default setting) for every datapoint that falls in that bin. We can also average on other directions, see "Set `on`" in Mapping system.

  * `any` reduction function:
  
```{r reduction on any, fig.width = 4, fig.height = 3, eval=FALSE}
r %>% 
  rasterize_points(
    reduction_func = "any",  # take the "any" reduction function 
    colour_map = c("white", "black")
  )
```

Availible reduction functions:

+ `sum`: If `on` is not provided in mapping `aes()`, it is the count of each bins; else it would be summation of 
`on` variable falling in that bin.

+ `any`: If `on` is not provided in mapping `aes()`, it is the logical value if the bin is visited; else it would be 
any `on` variable falling in that bin.

+ `mean`: If `on` is not provided in mapping `aes()`, `on` would be set as variable "y" by default. 
Give the mean of `on` variable falling in that bin.

+ `m2`: `on` must be provided in mapping `aes()`. Give the sum of mean squares of `on` variable falling in that bin.

+ `var`: `on` must be provided in mapping `aes()`. Give the variance of `on` variable falling in that bin.

+ `sd`: `on` must be provided in mapping `aes()`. Give the standard deviation of `on` variable falling in that bin.

+ `first`: `on` must be provided in mapping `aes()`. Give the first `on` variable element falling in that bin.

+ `last`: `on` must be provided in mapping `aes()`. Give the last `on` variable element falling in that bin.

+ `min`: `on` must be provided in mapping `aes()`. Give the max `on` variable element falling in that bin.

+ `max`: `on` must be provided in mapping `aes()`. Give the min `on` variable element falling in that bin.

## About

1. `rasterly` is only for large data set. There are so many good packages in R to display normal size data, like `grDevices`, `grid`, `ggplot2` for static display and `plotly`, `loon` for interactive display 

2. Package `rasterly` is sloppy right now, welcome to report issues in https://github.com/plotly/rasterly.

3. Further job: build `geom_rasterly()` for `ggplot2` and `l_rasterly()` for `loon` 